---
title: "Remote sensing classification of saltmarshes in the Ria Formosa"
subtitle: "A combination of traditional transects, remote sensing and machine learning"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

This project is a collection of the methods used during my work at mapping the saltmarsh communities in the Ria Formosa lagoon (South Portugal).

For this project, I took transects performed along saltmarsh fringes, where species presence/absence was recorded. This information was combined with the transect's start/end points to estimate a position for all quadrats along the transect. These quadrats were clustered into one of 3 possible saltmarsh communities.

To do so, unsupervised clustering methods were used to select saltmarsh areas. These areas were then classified into one of the 3 mentioned vegetation communites, using machine learning methods that relied on the transect data.

This was possible due to the availability of remote sensing data (3m resolution, R G B and NIR bands) as well as a LIDAR digital elevation model.

The project was fully implemented in open source, via a combination of R and QGIS.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
